Long name of the paper

### 2018-10-11 anzjs2018/sections/01_introduction.tex
type yp edits

From the comments posted on social media during outages,
it would seem that \rt information (RTI) has surpassed being a novelty
to become an integral component of public transport.
In Auckland, the RTI provided by Auckland Transport is unreliable at best,
and often succumbs to several major problems that public transport users
are all too familiar with.
The first is inaccurate estimated times of arrival (ETAs),
which tend to increase over time as the bus gets progressively later
due to traffic congestion, or simply because the schedule is poorly calibrated.
The second is buses that are shown as on time but in fact are not running,
and therefore have not registered on the \rt system in which case the default,
scheduled times are used as ETAs.

>>>>

% From the comments posted on social media during outages,
% it would seem that \rt information (RTI) has surpassed being a novelty
% to become an integral component of public transport.
In Auckland, the RTI provided by Auckland Transport is unreliable at best,
and often succumbs to several major problems that public transport users
are all too familiar with.
The first is inaccurate estimated times of arrival (ETAs),
which tend to either increase or decrease over time as the bus gets progressively later or earlier
due to traffic congestion, or simply because the schedule is poorly calibrated.
The second is buses that are shown as on time but in fact are not running,
and therefore have not registered on the \rt system in which case the 
scheduled times are used as ETAs.

### 2018-10-11 anzjs2018/sections/02_gtfs.tex
type yp edits

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data,
developed and maintained by Google \citep{GoogleDevelopers_2006}.
It is used by over 900~transit providers around the world,
including here in Auckland, New Zealand
(source \url{http://transitfeeds.com}),
allowing an application developed locally to
be deployed to another GTFS-based public transport system with minimal modification.

The two main components of GTFS are \emph{static} and \emph{realtime}.
Static GTFS includes information about routes, trips, shapes, and stops:
a \emph{route} is a sequence of two or more stops displayed as a single service;
a \emph{trip} is an instance of a route occuring at a specific time of day;
a \emph{shape} is the sequence of points defining a vehicle's path along a route;
and a \emph{stop} is a physical location where passengers can embark and disembark
the vehicle.
GTFS realtime feeds provide information such as vehicle positions and trip delays,
and are usually accessed via an API.
% The latter is often the only covariate used for predicting arrival time,
% as previously discussed:
% while it is a computationally inexpensive method (simply addition),
% it relies on accurate schedules and cannot respond quickly to realtime events.

@@

One solution would be to use information obtained from
vehicles servicing other routes but traveling along the same roads
to estimate arrival times;
however, there is no direct way of knowing which routes share common roads.
We are propose using an algorithm to convert the raw GTFS shape files
into a \emph{road network},
as demonstated in Figure~\ref{fig:network_creation}.
This involves detecting locations where one or more routes meet 
(intersections or \emph{nodes}),
and the connecting paths (road segments or \emph{edges}).

% Before we can usefully model buses in real-time, 
% we need to construct a \emph{transit network},
% consisting of \emph{intersections} (nodes)
% connected by \emph{road segments} (edges).
% In this way, each \emph{route} 
% (a journey taken by a transit vehicle, from an origin to a destination by a fixed path)
% can be represented as a sequence of road segments,
% each of which we model as vehicles travel along them (see section~\ref{sec:kf}).

@@

GTFS realtime allows developers to query the current positions of vehicles
in the transit network.
The data consists of the time $t_k$ that the observation was made,
the GPS position of the vehicle, $Y_k$, 
and some other information about the trip being serviced.
Vehicle positions are usually updated within 30~seconds,
but this can vary from 10~seconds to several minutes,
so there is often a lot of uncertainty about the trajectory
between two observations, particularly when there is a bus stop
or intersection between them.

Another complication with the Auckland Transport realtime feed is that
the buses are programmed to report their location when arriving at
bus stops and some major intersections.
Often these positions appear to be preemptive (i.e., that bus is almost there, 
but not quiet),
and subsequent observations place the bus \emph{behind} the stop or intersection.
This must therefore be accounted for in the model proposed in the next section.

>>>>

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data,
developed and maintained by Google \citep{GoogleDevelopers_2006}.
It is used by over 900~transit providers around the world,
including here in Auckland, New Zealand
(source \url{http://transitfeeds.com}),
allowing an application developed in Auckland to
be deployed to another GTFS-based public transport system with minimal modification.

The two main components of GTFS are \emph{static} and \emph{realtime}.
Static GTFS includes information about routes, trips, shapes, and stops:
a \emph{route} is a sequence of two or more stops displayed as a single service;
a \emph{trip} is an instance of a route occuring at a specific time of day;
a \emph{shape} is the sequence of points defining a vehicle's path along a route;
and a \emph{stop} is a physical location where passengers can embark and disembark
the vehicle \citep{GoogleDevelopers_2006}.
GTFS realtime feeds provide information such as vehicle positions and trip delays,
and are usually accessed via an API.
% The latter is often the only covariate used for predicting arrival time,
% as previously discussed:
% while it is a computationally inexpensive method (simply addition),
% it relies on accurate schedules and cannot respond quickly to realtime events.

@@

One solution would be to use information obtained from
vehicles servicing other routes but traveling along the same roads
to estimate arrival times;
however, there is no direct way of knowing which routes share common roads.
Here we propose using an algorithm to convert the raw GTFS shape files
into a \emph{road network},
as demonstated in Figure~\ref{fig:network_creation}.
This involves detecting locations where one or more routes meet 
(intersections or \emph{nodes}),
and the connecting paths (road segments or \emph{edges}).
However, in this paper we use bus stops as nodes and the routes between them as edges.
This captures most overlaps, 
with the exception of some express services and multiple-stop locations,
but these are typically in high traffic areas and should be OK.
Figure~\ref{fig:nw_construct} demonstrates how several overlapping routes 
are merged to form a road network.
In section~\ref{sec:kf} we present a model for estimating veihcle travel time
along a road segment.

@@

GTFS \rt allows developers to query the current positions of vehicles
in the transit network.
The data consists of the time $t_k$ that the observation was made,
the GPS position of the vehicle, $\bY_k$, 
and some other information about the trip being serviced.
Vehicle positions are updated with a frequency of anywhere from 10~seconds to several minutes,
so there is often a lot of uncertainty about the trajectory
between two observations, particularly when there is a bus stop
or intersection between them.

Another complication with the Auckland Transport realtime feed is that
the buses are programmed to report their location when arriving at
bus stops and some major intersections.
Often these positions appear to be preemptive (i.e., that bus is almost there, 
but not quite),
and subsequent observations place the bus \emph{behind} the stop or intersection.
To handle this, we compute the approximate distance traveled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the previous observation is rejected and the vehicle reverted
to its previous state.
To enable this, vehicles retain the current and previous states (see section~\ref{sec:rt}).

### 2018-10-11 anzjs2018/sections/03_models.tex
type yp edits

The real-time nature of this application (means that for) both
the vehicle and network models a recursive Bayesian filtering approach is logical,
and indeed real-time tracking and robotics applications often use such (models).
In each of the following models, 
we assume an underlying Markov process with state $\mathcal{X}_k$,
of which observations $\mathcal{Y}_k$ are made,
which leads us to the following general model,
\begin{equation}
\label{eq:rbe_model}
\begin{split}
\mathcal{X}_k &= f(\mathcal{X}_{k-1}, \omega_k) \\
\mathcal{Y}_k &= h(\mathcal{X}_k, \nu_k)
\end{split}
\end{equation}
where $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ is the system noise,
and $\nu_k\sim\mathrm{N}(0,\mathcal{R})$ is measurement error.
The transition function $f$ determines the relationship between consecutive states,
while the measurement function $h$ is a deterministic function describing the
relationship between the underlying and observed states.

The following sections describe two Bayesian filter models based on (\ref{eq:rbe_model}) used in this application.
The first model is implemented using a particle filter to estimate vehicle states,
with the primary objective of estimating vehicle travel times along roads,
while the second using these estimated times to update road states,
and is implemented using a Kalman filter.

@@

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance traveled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
Additionally, we include travel times along road segments along the route, 
$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
which are independent of time but included in the state as they are estimated
sequentially as the vehicle traverses the route.
These are combined into the state vector,
\begin{equation}
\label{eq:vehicle_state}
\bX_k = 
\begin{bmatrix}
    x_k \\ \dot x_k \\ \bz
\end{bmatrix}.
\end{equation}
Observations of the vehicle are made at time $t_k$ using a GPS,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle,
giving us the observation vector
\begin{equation}
\label{eq:vehicle_obs}
\bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
\end{equation}

We decided to employ a particle filter to estimate $\bX_k$,
since it is a highly flexible approach and has been used in recent 
transit vehicle modeling applications \citep{Hans_2015}.
Our main justification for using it is that it handles multimodality very well,
which is a common feature of the proposal distribution particularly around bus stops.
Another advantage is the intuitive likelihood function, 
which is described later.
Conversely, the particle filter is a computationally demanding method,
as thousands of particles are required per bus.
Section~\ref{sec:rt} describes how we were able to implement the particle filter in real-time.

In the particle filter, the posterior distribution of the state at time $t_k$,
is estimated by a set of discrete points, or particles,
\begin{equation}
p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := (\bX_k^{(i)})_{i=1}^N
\end{equation}
each of which is independently updated or \emph{mutated} using the transition function $f$,
\begin{equation}
p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
\left(f(\bX_{k-1}^{(i)}, \psi)\right)_{i=1}^N
\end{equation}
using a parameter vector $\psi$ containing all of the necessary parameters
for the model (including system noise).
After mutation, \emph{selection} of a new set of particles is performed by
importance resampling using likelihood weights.

\subsubsection{Vehicle transition function}
\label{sec:pf_prediction}
As we are using a particle filter to implement the vehicle model,
the transition function $f$ can be a complex description of bus behaviour,
most notably
\begin{itemize}
\item non-constant speed along roads (acceleration as system noise),
\item stopping and waiting at bus stops while passengers board and disembark, and
\item stopping and waiting at intersections.
\end{itemize}

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicles acceleration changes as a random process.
% In its simplest form, the transition function becomes,
% from Newton's laws of motions,
% \begin{equation}
% \begin{split}
%     x_k &= x_{k-1} + \delta_k \dot x_{k-1} \\
%     \dot x_k &= \dot x_{k-1} + w_k,
%     \quad w_k \sim \mathrm{N}(0, \delta_k \sigma^2)
% \end{split}
% \end{equation}
However, the vehicle does not simply travel constantly along the route,
as it needs to service bus stops along the way.
Therefore, the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$.
For these parameters, we used constant values for all stops,
and based the values on those used by \cite{Hans_2015};
future work will look at modeling these separately in real-time also.
Also in this application, due to the complexity of the problem,
we use bus stops to define road segments.
Therefore no intersection model is required,
only the segment travel times as predicted by the network model at time $t_k$,
$\btheta(t_k) = (\theta_1(t_k), \ldots, \theta_L(t_k))^\top$
(see section~\ref{sec:kf}).
Future work will adapt this to use intersection that are indpendent 
of bus stops, and therefore completely independent of routes,
which will allow increasing the complexity of the model to incorporate
intersection stopping probabilities and waiting times.
The details of the transition function are given in the algorithm defined in the Appendix,
which implements those features discussed above.

@@

After mutating the particle set, the posterior distribution is obtained by
\emph{selection}, which is implemented by importance resampling using
likelihood weights.
To do so, we need to define the measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection,
which allows us to place a bivate normal likelihood on the data.
The model for the observation generation is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
However, due to the assumption that error in longitude and latitude are uncorrelated, 
the geographical distance between the observation and the true vehicle can be expressed
in terms of two independent standard normal random variables $z_1$ and $z_2$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}
\end{equation}
As the sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

@@

The posterior distribution is obtained by \emph{selection},
or resampling with replacement using likelihood weights
\begin{equation*}
W_k^{(i)} = \frac{p(\bY_k | \bX_k^{(i)}, \epsilon)}{
    \sum_{j=1}^N p(\bY_k | \bX_k^{(j)}, \epsilon)
} 
W_{k-1}^{(i)} 
\end{equation*}
However, to improve efficiency, the state $\tilde \bX_k$ is only resampled
when the effective sample size 
$N_{\text{eff}}$ is less than some threshold,
\begin{equation*}
N_{\text{eff}} = \frac{1}{\sum_{i=1}^N (W_k^{(i)})} < N_{\text{thres}}
\end{equation*}
When resampling is performed, $W_k^{(i)} := N^{-1}$ for all $i$.

@@

The second part of this appraoch involves a network model of transit network travel times along links in the network.
Reliable ETAs will require a historical data component,
used as a prior in the absense of data as well as for
predicting future states, 
and a realtime update component where the travel times of vehicles
are used to update the current estimate of network state.

The model will be another recursive Bayesian model as used for the vehicle state,
as this provides a simple way to model realtime data.
Since travel times can safely be assumed to be unimodal,
we use an information filter (IF),
which is a transformation of the Kalman filter
using information instead of covariance,
and allows the simple combination of multiple observations,
for example when multiple buses traverse a road in the same update.
The prediction step will involve a transition function that incorporates historical data
and the update step will weight between observed and predicted travel times.

The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
of transit vehicles along road segment $\ell$ at time $t_c$.
In this implementation, we are assuming independence between each segment
to provide a simple basis for the model; of course in reality this
is not true and future work will investigate ways of incorporating this.

Since the road state model is very simple, 
no transition matrix or measurement matrix is required,
and the model from (\ref{eq:rbe_model}) reduces to
\begin{equation}
\begin{split}
\hat \theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
Z_c^\ell &= \theta_c^\ell + e_c^\ell
\end{split}
\end{equation}
where $\Delta_c = t_c - t_{c-1}$,
$v_c^\ell \sim \mathrm{N}(0, \nu^2)$ is the system noise,
and $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ is the measurement error
as estimated from the vehicle model using particle uncertainty.

@@

For the update step, however, we need to transform into an information
space parameterised by the information matrix $U^\ell_c$
and the information vector $u^\ell_c$,
\begin{equation}
\label{eq:information_transform}
U^\ell_{c|c-1} = \frac{1}{P_{c|c-1}}
\quad\text{and}\quad
\hat u^\ell_{c|c-1} = \frac{\hat \theta^\ell_{c|c-1}}{P^\ell_{c|c-1}}
\end{equation}
The measurement data obtained from the vehicle model are
the travel time of vehicle $m$ along segment $\ell$,
$\bar z_c^{m\ell}$, and the uncertainty $s^{m\ell}_c$.
These can be transformed to a measurement information covaraince matrix
and vector using the measurement matrix,
\begin{equation*}
B^{m\ell}_c = \frac{1}{(s^{m\ell}_c)^{2}}\quad\text{and}\quad
b^{m\ell}_c = \frac{\bar z^{m\ell}_c}{(s^{m\ell}_c)^2}
\end{equation*}
The information update is now the sum of the information for all vehicles
that traversed the segment since the last update, so
\begin{align*}
U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^M B^{m\ell}_{c} \\
\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^M b^{m\ell}_{c}
\end{align*}
which are back transformed to get the updated state parameters
using the inverse of (\ref{eq:information_transform})
\begin{equation*}
\hat \theta^\ell_{c|c} = \frac{\mathbf{\hat u}^\ell_{c|c}}{\mathbf{U}^\ell_{c|c}} 
\quad\text{and}\quad
P^\ell_{c|c} = \frac{1}{\mathbf{U}^\ell_{c|c}}.
\end{equation*}

>>>>

The real-time nature of this application suggests that
a recursive Bayesian filtering approach would be appropriate, as is often used in
real-time tracking and robotics applications.
We are using two separate models, one each for the vehicle and network states.
In each,
we assume an underlying Markov process with state $\mathcal{X}_k$,
of which observations $\mathcal{Y}_k$ are made,
giving us to the following general model,
\begin{equation}
\label{eq:rbe_model}
\begin{split}
\mathcal{X}_k &= f(\mathcal{X}_{k-1}, \omega_k) \\
\mathcal{Y}_k &= h(\mathcal{X}_k, \nu_k)
\end{split}
\end{equation}
where $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ is the system noise,
and $\nu_k\sim\mathrm{N}(0,\mathcal{R})$ is measurement error.
The transition function $f$ determines the relationship between consecutive states,
while the measurement function $h$ is a deterministic function describing the
relationship between the underlying and observed states.

The following sections describe the two Bayesian filter models based on (\ref{eq:rbe_model}) used in this application.
The first is implemented using a particle filter to estimate vehicle states,
with the primary objective of estimating vehicle travel times along roads,
while the second using these estimated times to update road states,
and is implemented using a variant of the Kalman filter referred to as the \emph{information filter}
\citep{}.

@@

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance traveled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
Additionally, we include travel times along road segments along the route, 
$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
which are independent of time but included in the state as they are estimated
sequentially as the vehicle traverses the route.
These are combined into the state vector,
\begin{equation}
\label{eq:vehicle_state}
\bX_k = 
\begin{bmatrix}
    x_k \\ \dot x_k \\ \bz
\end{bmatrix}.
\end{equation}
Observations of the vehicle are made using a GPS at time $t_k$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
and stored in the observation vector
\begin{equation}
\label{eq:vehicle_obs}
\bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
\end{equation}

We chose to employ a particle filter to estimate $\bX_k$,
to do its high flexibility and use in recent 
transit vehicle modeling applications \citep{Hans_2015}.
Our main justification for using it is that it handles multimodality very well,
as demonstrated in figure~\ref{fig:pf_state},
which is a common feature of the proposal distribution particularly around bus stops.
Another advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
Conversely, the particle filter is a computationally demanding method,
as thousands of particles are required per bus.
Section~\ref{sec:rt} describes how we were able to implement the particle filter in real-time and the timings.

In the particle filter, the posterior distribution of the state at time $t_{k-1}$,
is represented by a set of discrete points, or particles, (fig~\ref{fig:pf_state_prev})
\begin{equation}
p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} := (\bX_{k-1}^{(i)})_{i=1}^N
\end{equation}
each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
\begin{equation}
p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
\left(f(\bX_{k-1}^{(i)}, \psi)\right)_{i=1}^N
\end{equation}
using a parameter vector $\psi$ containing all of the necessary parameters
for the model (including system noise).
After mutation, \emph{selection} of a new set of particles is performed by
importance resampling using likelihood weights,
determined by distance between particle's map position and $\bY_k$ (fig~\ref{fig:pf_state_map})
to obtain a new sample $\tilde\bX_{k|k}$, (fig~\ref{fig:pf_state_update}).

\subsubsection{Vehicle transition function}
\label{sec:pf_prediction}

Another advantage of using a particle filter to implement the vehicle model is that
the transition function $f$ can implement several bus behaviours,
most importantly
\begin{itemize}
\item non-constant speed along roads (acceleration as system noise),
\item optional stopping and waiting at bus stops while passengers board and disembark, and
\item optional stopping and waiting at intersections.
\end{itemize}

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicles acceleration changes as a random process.
However, the vehicle does not travel constantly along the route,
as it needs to service bus stops along the way.
Therefore, the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$.
For these parameters, we used constant values for all stops,
and based the values on those used by \cite{Hans_2015};
future work will look at modeling these separately in real-time similarly to \ref{sec:kf}.
As we are using bus stops to determined segments,
no intersection model is required,
only the segment travel times as predicted by the network model at time $t_k$,
$\btheta(t_k) = (\theta_1(t_k), \ldots, \theta_L(t_k))^\top$
(see section~\ref{sec:kf}).
Future work will adapt this to use intersection that are indpendent 
of bus stops, and therefore completely independent of routes,
which will allow increasing the complexity of the model to incorporate
intersection stopping probabilities and waiting times.
The details of the transition function are given in the algorithm defined in the Appendix,
which implements those features discussed above.

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ is obtained by
\emph{selection}, which is implemented by importance resampling using
likelihood weights.
To do so, we define the measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection),
which allows us to place a bivate normal likelihood on the data
as shown in figure~\ref{fig:pf_lhood},
The model for the observation generation is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
However, due to the assumption that error in longitude and latitude are uncorrelated, 
the geographical distance between the observation and the true vehicle can be expressed
in terms of two independent standard normal random variables $z_1$ and $z_2$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}
\end{equation}
As the sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

@@

The posterior distribution is obtained by \emph{selection},
or resampling with replacement using the likelihood weights
\begin{equation*}
W_k^{(i)} = \frac{p(\bY_k | \bX_k^{(i)}, \epsilon)}{
    \sum_{j=1}^N p(\bY_k | \bX_k^{(j)}, \epsilon)
} 
W_{k-1}^{(i)} 
\end{equation*}
However, to improve efficiency, the state $\tilde \bX_k$ is only resampled
when the effective sample size 
$N_{\text{eff}}$ is less than some threshold,
\begin{equation*}
N_{\text{eff}} = \frac{1}{\sum_{i=1}^N (W_k^{(i)})} < N_{\text{thres}}
\end{equation*}
When resampling is performed, $W_k^{(i)} := N^{-1}$ for all $i$.
When sampling is not performed,
state estimates use $\{\bX_k^{(i)}, W_k^{(i)}\}$ to obtain weighted estimates.

@@

[[  blurb  ]]
% The second part of this appraoch involves a network model of transit network travel times along links in the network.
% Reliable ETAs will require a historical data component,
% used as a prior in the absense of data as well as for
% predicting future states, 
% and a realtime update component where the travel times of vehicles
% are used to update the current estimate of network state.

To obtain reliable ETAs,
it is necessary to estimate travel times along roads,
as this allows the model to react to \rt congestion as detected by other buses.
Future work will aim to construct a more complex model incorporating historical trends 
and correlations between roads, 
however for the current work we are demonstrating our approach
using a simple model.

The model assums a Gaussian state distribution for travel time,
truncated necessarily at zero;
this will not be an issue as the state prediction and data will always be non-negative.
However, the model must be able to handle multiple observations in a single update,
which often occurs along major bus routes when several buses are traveling together.
Therefore we have chosen the information filter as our estimation method.

The information filter is a transormation of the Kalman filter in which the
\emph{information matrix} and \emph{information vector} are used in place of 
the covariance matrix and state vector, respectively.
The advantage is that multiple observations can be added together when updating the state.
Otherwise the information filter follows the same procedure as other recursive Bayesian models.

The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
of transit vehicles along road segment $\ell$ at time $t_c$.
Observations of the state, $Z_c^\ell$, are obtained from the particle filter above.
We are assuming independence between each segment
to provide a simple basis for the model; of course in reality this
is not true and future work will investigate ways of incorporating this.
Thanks to the model simplicity,
no transition matrix or measurement matrix is required,
and the model from (\ref{eq:rbe_model}) reduces to
\begin{equation}
\begin{split}
\hat \theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
Z_c^\ell &= \theta_c^\ell + e_c^\ell
\end{split}
\end{equation}
where $\Delta_c = t_c - t_{c-1}$,
$v_c^\ell \sim \mathrm{N}(0, \nu^2)$ is the system noise,
and $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ is the measurement error
as estimated from the vehicle model using particle uncertainty.

@@

For the update step, the parameters are transformed into an information
space parameterised by the information matrix $U^\ell_c$
and the information vector $u^\ell_c$,
\begin{equation}
\label{eq:information_transform}
U^\ell_{c|c-1} = \frac{1}{P_{c|c-1}}
\quad\text{and}\quad
\hat u^\ell_{c|c-1} = \frac{\hat \theta^\ell_{c|c-1}}{P^\ell_{c|c-1}}
\end{equation}
The measurement data obtained from the vehicle model are
the travel time of vehicle $m$ along segment $\ell$,
$\bar z_c^{m\ell}$, and the uncertainty $s^{m\ell}_c$.
These are transformed to a measurement information covariance matrix $B$
and vector $b$ using
\begin{equation*}
B^{m\ell}_c = \frac{1}{(s^{m\ell}_c)^{2}}\quad\text{and}\quad
b^{m\ell}_c = \frac{\bar z^{m\ell}_c}{(s^{m\ell}_c)^2}
\end{equation*}
The information update is now the sum of the information for all vehicles
that traversed the segment since the last update, so
\begin{align*}
U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^M B^{m\ell}_{c} \\
\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^M b^{m\ell}_{c}
\end{align*}
which are back transformed to get the updated state parameters
using the inverse of (\ref{eq:information_transform})
\begin{equation*}
\hat \theta^\ell_{c|c} = \frac{\hat u^\ell_{c|c}}{U^\ell_{c|c}} 
\quad\text{and}\quad
P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}.
\end{equation*}
